---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2025"
author: "Daniel Whitehead"
subtitle: "Assignment 2 - Due date 01/23/25"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
geometry: margin=2.54cm
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.   

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp24.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(forecast)
library(tseries)
library(dplyr)
library(ggplot2)
```


## Data set information  

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our 
**Data** folder. The data comes from the US Energy Information and 
Administration and corresponds to the December 2023 Monthly Energy Review. 
The spreadsheet is ready to be used. You will also find a $.csv$ version of the 
data "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source-Edit.csv". 
You may use the function $read.table()$ to import the $.csv$ data in R. Or refer
to the file "M2_ImportingData_CSV_XLSX.Rmd" in our Lessons folder for functions 
that are better suited for importing the $.xlsx$.

```{r}
#Importing data set
library(readxl)
Table_10_1_Renewable_Energy_Production_and_Consumption_by_Source <- read_excel(
  "~/ENV797/Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx", 
  skip = 9)
Energy_df <- 
  Table_10_1_Renewable_Energy_Production_and_Consumption_by_Source[-1, ]

print(Energy_df)
```


## Question 1
You will work only with the following columns: Total Biomass Energy Production, 
Total Renewable Energy Production, Hydroelectric Power Consumption. Create a 
data frame structure with these three time series only. Use the command head() 
to verify your data.

```{r Question 1}

selected_energy <- Energy_df %>%
  select(4,5,6)
#realized my data was in character form :(
selected_energy <- as.data.frame(lapply(selected_energy, as.numeric))

head(selected_energy)
```


## Question 2
Transform your data frame in a time series object and specify the starting 
point and frequency of the time series using the function ts().

```{r Question 2}
energy_ts <- ts(selected_energy, start = c(1973, 1), frequency = 12)

head(energy_ts)
```


## Question 3
Compute mean and standard deviation for these three series.

```{r Mean and Standard Deviation}
#Biomass
mean_biomass <- round(mean(energy_ts[, 1]), 2)
sd_biomass <- round(sd(energy_ts[, 1]), 2)

# Renewable 
mean_renewable <- round(mean(energy_ts[, 2]), 2)
sd_renewable <- round(sd(energy_ts[, 2]), 2)

# Hydroelectric 
mean_hydroelectric <- round(mean(energy_ts[, 3]), 2)
sd_hydroelectric <- round(sd(energy_ts[, 3]), 2)

print(paste("Biomass Mean and Standard Deviation:"))
print(paste("Mean:", mean_biomass, "trillion BTU. Standard Deviation:", 
            sd_biomass, "trillion BTU."))

print(paste("Renewable Mean and Standard Deviation:"))
print(paste("Mean:", mean_renewable, "trillion BTU. Standard Deviation:", 
            sd_renewable, "trillion BTU."))

print(paste("Hydroelectric Mean and Standard Deviation:"))
print(paste("Mean:", mean_hydroelectric, "trillion BTU. Standard Deviation:", 
            sd_hydroelectric, "trillion BTU."))

```


## Question 4
Display and interpret the time series plot for each of these variables. Try to 
make your plot as informative as possible by writing titles, labels, etc. For 
each plot add a horizontal line at the mean of each series in a different color.


```{r Combined Plot}

autoplot(energy_ts) +
  labs(
    x = "Year",  
    y = "Energy Production/Consumption (Trillion BTU)", 
    title = "Energy Production/Consumption Over Time", 
    caption = "Figure 1: Time series of energy production and consumption, 
    with dashed lines indicating the mean values for Total 
    Biomass Energy Production (red), Total Renewable Energy 
    Production (green), and Hydroelectric Power Consumption (blue)."  
  ) +
  geom_hline(yintercept = mean_biomass, color = "red", linetype = "dashed", 
             size = 1.2) +
  geom_hline(yintercept = mean_renewable, color = "green", linetype = "dashed", 
             size = 1.2) +
  geom_hline(yintercept = mean_hydroelectric, color = "blue", 
             linetype = "dashed", size = 1.2) +
  theme(plot.caption = element_text(hjust = 0.5, size = 10, 
                                    face = "italic")) 

```
For total renewable energy production, production is at about 200 trillion BTU 
in 1973. It increases for about 15 years, stagnates for about 20, then hits a 
(somewhat steep) steady increase that continues to grow into the 2020's. Biomass 
has a very similar trend, though it starts closer to 125 trillion BTU in 1973 
and stagnates a second time starting in about 2015. Hydroelectric consumption 
remains steady abound 80 trillion BTU throughout the entirety of the time series 
plot.



## Question 5
Compute the correlation between these three series. Are they significantly
correlated? Explain your answer.

```{r Correlation}

correlation_matrix <- cor(energy_ts, use = "complete.obs")

print("Correlation Matrix:")
print(correlation_matrix)

```
Total Biomass Energy Production and Total Renewable Energy Production seem to be
highly correlated as there is a 0.97 correlation between them. However, these 
series do not seem to have much correlation with Hydroelectric Power 
Consumption, as the correlation values are very low and insignificant.



## Question 6
Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

```{r}

acf_biomass <- Acf(energy_ts[,1], lag.max = 40, 
                   main = "ACF of Total Biomass Energy Production", 
                   type = "correlation", plot = TRUE)
acf_renewable <- Acf(energy_ts[,2], lag.max = 40, 
                     main = "ACF of Total Renewable Energy Production",
                     type = "correlation", plot = TRUE)
acf_hydro <- Acf(energy_ts[,3], lag.max = 40, 
                 main = "ACF of Hydroelectric Power Consumption", 
                 type = "correlation", plot = TRUE)
```
Total Biomass Energy Production and Total Renewable Energy Production have 
similar ACF plots in terms of behavior. Both start with high values and then 
slowly decline. Hydroelectric Power Consumption is different in terms of 
behavior, as it rises and falls, crossing in and out of negative. values. This
means that for Biomass and Renewables, there is more predictability in 
forecasting the future or longer-term data. Hydro seems to be more affected by 
factors that cause short term fluctuations.




## Question 7
Compute the partial autocorrelation function from lag 1 to lag 40 for these 
three variables. How these plots differ from the ones in Q6?

```{r Partial ACF version}
pacf_biomass <- Pacf(energy_ts[,1], lag.max = 40, 
                     main = "PACF of Total Biomass Energy Production", 
                     plot=TRUE)
pacf_renewable <- Pacf(energy_ts[,2], lag.max = 40, 
                       main = "PACF of Total Renewable Energy Production", 
                       plot=TRUE)
pacf_hydro <- Pacf(energy_ts[,3], lag.max = 40, 
                   main = "PACF of Hydroelectric Power Consumption", plot=TRUE)
```
These plots differ because they have much shorter bars. This makes sense because
in general the PACF plots have shorter bars, as they do not account for the 
complete correlation of the lags and instead do not account for shorter lags as 
the lags get higher in values. Additionally, the hydropower plot is much more 
similar to the biomass and renewable plots.


